---
title: "Notebook 07"
subtitle: "DSST289: Introduction to Data Science"
date: "`r Sys.Date()`"
format:
  html:
    embed-resources: true
    link-external-newwindow: true
    df-print: paged
    smooth-scroll: true
    toc: true
editor_options:
  markdown:
    wrap: 72
  chunk_output_type: inline
---


## Getting Started

Before running this notebook, select **Session > Restart R and Clear Output** in
the menu above to start a new R session. This will clear any old data sets and
give us a blank slate to start with.

After starting a new session, run the following code chunk to load the
libraries and data that we will be working with today.

```{r, include=FALSE, message=FALSE}
library(tidyverse)
library(ggrepel)

if (!require(ggthemes)) {
  install.packages("ggthemes")
  library(ggthemes)
} else {
  library(ggthemes)
}
```

I have set the options `include=FALSE` and `message=FALSE` to avoid cluttering
the solutions with all the output from this code.

## Practice

### DIY movie data

For this notebook, we will be creating a dataset, reading it into R, and
producing a couple of different data visualizations.

We will create a dataset as a class describing movies that we have seen recently.
Each student will be tasked with creating **five** rows of data. We will record data in the Google Sheets file linked to on Blackboard and [here](https://docs.google.com/spreadsheets/d/189QVmv5_2raf_ii-49UKnud4n3ITPmRcKOdHZzNs0G8/edit?usp=sharing).

### Data collection process

It is a good idea to collect data in a structured (and, ideally, reproducible) way.
In the Google Sheet above, we're going to collect the data in the columns
as defined in the `dictionary` tab.
Categorical variables (like genres) will be filled in using 
values from the `validation` tab.

The principle of selection for our observations is as follows:

Everyone will enter data for **five movies you remember seeing**. This could be
a movie you saw last night, or it could be a memorable experience in childhood.
Think of five memorable movie-watching experiences, and collect data about those
movies.

Most information about our memorable movies will come directly from [IMDB](https://www.imdb.com/), but some will also capture our class's tastes and preferences.

## `read_csv` movie data

Once you are finished with your entries, save the current version of the data set as a `.csv`.

In Google Sheets, do this by going to **File > Download > Comma Separated Values (.csv)**

This should download a file called `movies - data.csv` to your computer. The
general pattern for Google Sheets is `filename - sheet.csv`, where `filename` is the name of the workbook
in Google Sheets, and `sheet` is the name of the sheet that you *are viewing*
when you download the `.csv`

Move that file to your `DSST289/data` directory.

Then, read it into R with the following block of code:

```{r, message=FALSE}
movies <- read_csv("../data/movies - data.csv")

movies |>
  arrange(year) |>
  select(title, year, imdb_rating)
```

Then, produce a scatterplot below with the budget on the x-axis, gross worldwide
earnings on the y-axis, points colored by the year of release, and the size of
the points corresponding to contributors' ratings. Use a text repel layer to
label the movies by their title. Try using `scale_color_viridis_b()` to color
the points in **bins** (hence `_b`). If you think it is needed, you may also
include logarithmic scales for the x- and y-axes

```{r, warning=FALSE, q1}

```

### `mutate` review

Let's use the data about the length of our films to create a new variable that
describes the **total** length in minutes. This is the kind of thing `mutate` is
great for:

```{r, q2}

```

Now, let's see if there appears to be any relationship between budget and length.
Plot `budget` on the x-axis and `total_minutes` on y. As before, color points by
`year`. Label the films and use viridis.

```{r, q3}

```


### Class rating vs. IMDB rating

Let's compare the ratings that we have given to the movies with IMDB ratings for
the movies.

First, since there are probably going to be some overlapping movies, we
might want to calculate the `mean` of our reviews for each movie.
We also need the IMDB rating in the resulting table.

We can group our results by title, and summarize the `contributor_rating`s.
All of the IMDB ratings *should* be the same, so we can use the `first()`
summary function, which will take the first value in a column.^[If we are not certain
that our data is perfect, this may not be the best choice! It would be smarter
to check and see that all values in the `imdb_rating` column are the same before
taking the `first()`.]

We can also size the points based on the ratio of the `mean_class_rating` to the
`imdb_rating`. If the class gave the film a higher review than IMDB, the value will be
greater than 1. If the class gave the film a lower review than IMDB, the value will be
less than 1.

```{r, q4}

```


### Labels and Themes

For the last few questions, let's finish a few loose ends about how to make
data visualizations in R. The function `labs()` is an important element that
can be added to any ggplot graphic. Inside the `labs()` function, you can
set aesthetics from the plot equal to a string to change the corresponding
label in the plot. For example, adding `labs(x = "HELLO!")` will change the
label of the x-axis to the string "HELLO!" In the code below, take the
previous code from above and add an appropriate label to the x-axis.

```{r, q5}

```

Now, copy the code above and in the next block add appropriate labels for
the y-axis, the color aesthetic, and the size aesthetic.

```{r, q6}

```

There are three special additional elements that we can set inside of 
`labs()` called `title` (the main title of the plot), `subtitle`,
and `caption` (the description at the bottom of the plot). Take the
code above and add a title, subtitle, and caption.

```{r, q7}

```

As a final step, we can add another element called a theme to change
the additional visual elements of the plot, such as what fonts are 
used in the labels, and what colors are used in the interior parts
of the graphic. The default that I have set for you in class is
called `theme_minimal()`. It's great for a first-pass of some graphics.

In the plot below, copy your graphic from above and add one final
element called `theme_tufte()`. This theme is named after [Edward Tufte](https://en.wikipedia.org/wiki/Edward_Tufte), who has been highly influential
in data visualization. Among the many differences between his style and
what we have seen so far, Tufte uses serif rather than sans-serif fonts
in his graphics. We can change the font `family` used by default (Helvetica)
to match this style with Times New Roman in `geom_text_repel()`.
We can do the same by setting the `base_family` argument in `theme_tufte()` to
`"Times New Roman"`.

```{r, q8}

```

I like this theme a lot for creating graphics for publication because 
it is very clean and avoids using extra ink that I can make better use
of with manual annotations. There are other themes that you can read about in 
the R documentation to change the way a `ggplot` object looks.
Several other R packages also provide additional
themes ready to use out of the box. 
