---
title: "Introduction to grammar of graphics"
author: "Erik Fredner"
date: "`r Sys.Date()`"
format:
  revealjs:
    #incremental: true
    theme: default
    transition: slide
    logo: "images/by-sa.png"
    footer: "https://fredner.org"
    embed-resources: true
---

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
```

# What is the grammar of graphics?

## Origin

- `ggplot2` did not create "the grammar of graphics."
- `ggplot2` implements the ideas in [*The Grammar of Graphics* (2005) by Leland Wilkinson](https://richmond.primo.exlibrisgroup.com/permalink/01URICH_INST/10lhjt5/cdi_springer_books_10_1007_0_387_28695_0).

## Key concepts

- Data visualizations (simple or complex) are composed of layers.
- Each layer consists of three parts:

| **Key**  | **Description**                                                                                     |
|----------|-----------------------------------------------------------------------------------------------------|
| `data`   | Tabular dataset associated with the layer                                                           |
| `geom`   | Graphical element associated with each observation                                                  |
| `aes`    | Mappings from properties of the plot that associate features in the dataset with elements of the geometry |

## Example: scatter plot of `food`

```{r, echo=TRUE}
food <- read_csv("../data/food.csv")

food |>
  ggplot(aes(x = calories, y = carbs)) +
  geom_point()
```

TODO figure out how to make the plot smaller
