---
title: "Aesthetics & Scales"
author: "Erik Fredner"
date: "`r Sys.Date()`"
echo: true
format:
  revealjs:
    #incremental: true
    theme: default
    transition: slide
    logo: "images/by-sa.png"
    footer: "https://fredner.org"
    embed-resources: true
    scrollable: true
    code-fold: true
---

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(ggrepel)
```

## Aesthetics: so what?

- Aesthetics (such as color, size, shape, etc.) determine how data points are visually distinguished in a plot.
  - Choosing the right aesthetics ensures that the visualization communicates the correct message.
  - For example, it would be confusing to color Democrats red and Republicans blue in a political visualization.

## Scales: so what?

- Scales control how data is mapped onto visual dimensions like the x- and y-axes.
  - This affects how easily readers can interpret the visualization.
  - Proper scaling can prevent misleading representations.
  
## Aesthetics & Scales with Pokémon

This shows the Pokemon with the highest `defense` and `hp` in the top-right corner of the plot.

```{r}
pokemon <- read_csv("../data/pokemon.csv")

pokemon |>
  ggplot() +
  geom_point(aes(x = defense, y = hp))
```

## Modifying scales

Let's suppose we wanted to flip that and see the Pokemon with the highest `defense` and **lowest** `hp` in the top-right corner.

```{r}
pokemon |>
  ggplot() +
  geom_point(aes(x = defense, y = hp)) +
  # reverse the y-axis
  scale_y_reverse()
```

## Combining scales and aesthetics

Let's find out the name of the Pokemon with low `hp` and high `defense`:

```{r}
pokemon |>
  ggplot() +
  geom_point(aes(x = defense, y = hp)) +
  scale_y_reverse() +
  # new:
  geom_text(aes(x = defense, y = hp, label = name))
```

## Limiting scales

```{r}
pokemon |>
  ggplot() +
  geom_point(aes(x = defense, y = hp)) +
  scale_y_reverse() +
  # repel the text labels:
  geom_text_repel(aes(x = defense, y = hp, label = name)) +
  # limit the x-axis to `defense` of 150 or more:
  # `NA` ("Not Available") is a missing value indicator.
  # We use it here to say that there is no upper limit on the x-axis.
  scale_x_continuous(limits = c(150, NA))
```

## Increasing `n.breaks`

```{r}
pokemon |>
  ggplot() +
  geom_point(aes(x = defense, y = hp)) +
  scale_y_reverse() +
  geom_text_repel(aes(x = defense, y = hp, label = name)) +
  # make it easier to identify the precise values of `defense`:
  scale_x_continuous(limits = c(150, NA), n.breaks = 30)
```

## Color

- We can use color to see patterns in the data by factors.
- Let's see if there are any patterns in the `type_1` of the Pokemon in relation to their `defense` and `hp`.
- And we're going to `filter` for first generation Pokemon for simplicity.

## Color by `type_1`

```{r}
pokemon |>
  filter(generation == 1) |>
  ggplot() +
  geom_point(aes(x = defense, y = hp, color = type_1)) +
  geom_text_repel(aes(x = defense, y = hp, label = name))
```

## Custom color

Let's use colors associated with 🔥, 🍃, and 💧 Pokemon:

```{r}
pokemon |>
  filter(generation == 1) |>
  # filter for just a few types
  filter(type_1 %in% c("Water", "Fire", "Grass")) |>
  ggplot() +
  geom_point(aes(x = defense, y = hp, color = type_1)) +
  geom_text_repel(aes(x = defense, y = hp, label = name)) +
  # use the `type_1` colors instead of the default:
  scale_color_manual(values = c(
    Water = "blue",
    Fire = "red",
    Grass = "green"
  ))
```

## `scale_color`

`stat_total` roughly represents how powerful a Pokemon is, which makes it obvious how good Mewtwo is.

```{r}
pokemon |>
  filter(generation == 1) |>
  ggplot() +
  # color the points by `stat_total` instead of `type1`:
  geom_point(aes(x = defense, y = hp, color = stat_total)) +
  # use the `viridis` color palette instead of the default:
  scale_color_viridis_c() +
  geom_text_repel(aes(x = defense, y = hp, label = name))
```

## Size

We could also use size to represent `stat_total`, which makes it obvious how bad Magikarp is!

```{r}
pokemon |>
  filter(generation == 1) |>
  # just water pokemon
  filter(type_1 == "Water") |>
  ggplot() +
  geom_point(aes(x = defense, y = hp, size = stat_total)) +
  # this scales the difference in size between points:
  scale_size_continuous(range = c(1, 10)) +
  geom_text_repel(aes(x = defense, y = hp, label = name))
```

## Summary

- **Aesthetics** determine how data points are visually distinguished, including aspects like color, size, and shape.
- **Scales** control how data is mapped onto visual dimensions such as x- and y-axes. Proper scaling ensures that visualizations are easy to interpret and not misleading.
- Manipulating both aesthetics and scales can reveal patterns and/or outliers in data.
