---
title: "Organizing *Star Wars* data"
subtitle: "DSST 289: Introduction to Data Science"
author: "Erik Fredner"
date: "`r Sys.Date()`"
echo: true
format:
  revealjs:
    logo: "images/by-sa.png"
    footer: "https://fredner.org"
    embed-resources: true
    scrollable: true
    code-fold: true
    tables: true
---

# Verbs 

## in language

- In spoken language, a *verb* is:

> 1.a. Grammar. A word or lexical unit which is used to indicate the occurrence or performance of an action or the existence of a state or condition, and which generally forms the grammatical predicate of the subject, or functions as the predicate of the subject in combination with an adjunct or adjuncts. ([Oxford English Dictionary](https://doi.org/10.1093/OED/3440824690))

## in R

- In R, we refer to some `dplyr` functions as *verbs*.
- Generally, these verbs manipulate data frames to achieve some goal.
- It's pronounced "d plier," as in "data plier, because just as you can use
pliers to manipulate physical objects, so too can you use `dplyr` to manipulate data.

## Single- vs. multi-table

- Today's notes and slides focused on single-table verbs.
- We will also discuss multi-table verbs, which are used to combine data from
multiple tables.
  - Combining data from several places is one of the most common tasks
  in data science.
  
# Applying verbs to a single `tibble`

## `tibble`: *Star Wars* characters

```{r}
#| code-fold: false
library(tidyverse)
# we're looking at a lot of tables today
# and the `kable` function from `knitr` will make them look nicer:
library(knitr)

starwars <- read_csv("../data/starwars.csv")
```

## `select`

With `select`, we can choose specific columns:

```{r}
starwars |>
  # let's see the names and heights of the characters
  select(name, height) |>
  # pretty-print the table:
  kable()
```

## `slice`

`slice` is used to select rows by position:

```{r}
starwars |>
  select(name, height) |>
  # get rows 10 through 13:
  slice(10:13) |>
  kable()
```

## `slice_head` and `slice_tail`

Often you want to see the top (head) or bottom (tail) of a `tibble`:

```{r}
starwars |>
  select(name, height) |>
  # let's see the last 5 rows, the `tail`:
  slice_tail(n = 5) |>
  kable()
```


## `filter`

We can select rows based on values in one or more columns using `filter`:

```{r}
starwars |>
  # let's see who isn't human:
  filter(species != "Human") |>
  select(name, species) |>
  kable()
```

## `filter` multiple conditions

You can `filter` on multiple conditions:

```{r}
#| code-fold: false
#| code-line-numbers: "3"
starwars |>
  # who isn't human and is also short?
  filter(species != "Human", height < 75) |>
  select(name, species, height) |>
  kable()
```

## `filter` and `%in%`

Use `%in%` to filter based on multiple values in a vector:

```{r}
#| code-line-numbers: "3"
starwars |>
  # who's a droid or a Wookiee?
  filter(species %in% c("Droid", "Wookiee")) |>
  select(name, species) |>
  kable()
```


## not (`!`) `%in%`

You can also invert it with `!`:

```{r}
#| code-line-numbers: "3"
starwars |>
  # which characters' sexes are neither male nor female?
  filter(!sex %in% c("male", "female")) |>
  select(name, sex) |>
  kable()
```


## `arrange`

`arrange` reorders rows based on the values in one or more columns.

```{r}
#| code-line-numbers: "3"
starwars |>
  # let's see who's the shortest
  arrange(height) |>
  select(name, height) |>
  slice_head(n = 3) |>
  kable()
```

## `arrange` multiple conditions

Like `filter`, `arrange` can handle multiple conditions:

```{r}
#| code-line-numbers: "3"
starwars |>
  # who's the lightest and the shortest?
  arrange(mass, height) |>
  select(name, mass, height) |>
  slice_head(n = 3) |>
  kable()
```

## `desc()` sorts in descending order

```{r}
#| code-line-numbers: "3"
starwars |>
  # who's the heaviest and the tallest?
  arrange(desc(mass), desc(height)) |>
  select(name, mass, height) |>
  slice_head(n = 3) |>
  kable()
```

## `fct_inorder()` orders factors for plots

```{r}
# let's see what the most frequent hair colors are:
starwars |>
  # fct_infreq orders by frequency:
  ggplot(aes(y = fct_infreq(hair_color))) +
  # Note that `NA` values are not ordered in the same way as other values
  geom_bar() +
  labs(y = "Hair color", x = "Count")
```

# Summary

We explored how `dplyr` verbs manipulate data, similar to how verbs in language describe actions.

## Single-Table Verbs

- `select`: Extract specific columns.
- `slice`: Select rows by position.
- `slice_head` & `slice_tail`: Get the top or bottom rows.
- `filter`: Choose rows based on column values.
- `arrange`: Reorder rows by column values.

## Key Functions

- `%in%`: Filter based on one or more values in a vector `c()`.
- `desc()`: Sort in descending order.
- `fct_infreq()`: Order factors by frequency for plotting.