---
title: "Tidy Models and Normal Forms"
subtitle: "DSST 289: Introduction to Data Science"
author: "Erik Fredner"
date: today
echo: true
format:
  revealjs:
    logo: "images/by-sa.png"
    footer: "https://fredner.org"
    embed-resources: true
    scrollable: true
editor_options:
  markdown:
    wrap: 72
---

# Homework review

```{r}
#| echo: false

library(broom)
library(ggrepel)
library(glue)
library(tidyverse)
theme_set(theme_minimal())
```

## Data

```{r}
hans <- read_csv("../data/hans_roslin.csv")

hans_subset <- hans |>
  filter(year == 2007) |>
  filter(continent == "Americas") |>
  select(country, year, gdp, life_exp)

hans_subset |>
  slice_sample(n = 5)
```

## Linear model

```{r}
model <- lm(life_exp ~ gdp, data = hans_subset)
```

- Read tilde (`~`) as "is modeled by."
- Life expectancy is modeled by GDP.
- Independent variable: GDP
- Dependent variable: Life expectancy

## `augment()`

```{r}
model <- lm(life_exp ~ gdp, data = hans_subset)

model |>
  augment(newdata = hans_subset)
```


## Plot

```{r}
#| output-location: slide
#| code-line-numbers: "|1|4|8|"
model <- lm(life_exp ~ gdp, data = hans_subset)

model |>
  augment(newdata = hans_subset) |>
  ggplot(aes(gdp, life_exp)) +
  geom_point() +
  geom_text_repel(aes(label = country)) +
  geom_line(aes(y = .fitted), color = "red", linetype = "dashed")
```

## 1. Output of `tidy()`

```{r}
model |>
  tidy()
```

## 2. Output of `augment()`

```{r}
model |>
  augment(newdata = hans_subset) |>
  slice_sample(n = 5)
```

## 3. Residuals

```{r}
#| output-location: slide
#| code-line-numbers: "1|2|3|5|"
model |>
  augment(newdata = hans_subset) |>
  ggplot(aes(gdp, .resid)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  geom_text_repel(aes(label = country))
```

## 3. Positive residuals

```{r}
model |>
  augment(newdata = hans_subset) |>
  arrange(desc(.resid)) |>
  slice_head(n = 5)
```

The largest positive residuals indicate which countries have the highest life expectancy compared to the model's prediction.

## 3. Negative residuals

```{r}
#| code-line-numbers: "1-4|5-10"
model |>
  augment(newdata = hans_subset) |>
  arrange(.resid) |>
  slice_head(n = 5) |>
  # highlight the US with flag emoji:
  mutate(country = if_else(
    condition = country == "United States",
    true = glue("🇺🇸 {country} 🇺🇸"),
    false = country
  ))
```

The largest negative residuals indicate which countries have the lowest life expectancy compared to the model's prediction.

# Data normalization

## Overview

- Data normalization is a best practice: Ideally, every *fact* in a data set should be stored in one and only one place.
  - Repeating facts in multiple locations wastes space, increases the chance of error, and makes it harder to update information.
- **Objective**: Organize data to minimize redundancy and dependency
- Now that we can `join` and `pivot`, we can benefit from normalization

## First Normal Form (1NF)

| Year | Tournament         | Tournament Country | Winner           | Winner Nationality |
|------|--------------------|--------------------|------------------|--------------------|
| 2020 | Australian Open    | AUS                | Sofia Kenin      | USA                |
| 2020 | French Open        | FRA                | Iga Świątek      | POL                |
| 2020 | Wimbledon          | GBR                | Naomi Osaka      | JPN                |
| 2020 | US Open            | USA                | Naomi Osaka      | JPN                |
| 2021 | Australian Open    | AUS                | Naomi Osaka      | JPN                |
| 2021 | French Open        | FRA                | Barbora Krejčíková | CZE              |
| 2021 | Wimbledon          | GBR                | Ashleigh Barty   | AUS                |
| 2021 | US Open            | USA                | Emma Raducanu    | GBR                |

## Primary key

In this case, the primary key is the combination of Tournament and Year. We **must** use both to uniquely identify a row.

## Problems with 1NF?

- *Repeated data*: Tournament Country and Winner Nationality
- Repeating data wastes space
- More importantly, if/when data changes, repeated data *must be changed in multiple places*

## Second Normal Form (2NF)

**Table 1: Tournament Information**

*Note the absence of winner nationality.*

| Year | Tournament       | Tournament Country | Winner           |
|------|------------------|--------------------|------------------|
| 2020 | Australian Open  | AUS                | Sofia Kenin      |
| 2020 | French Open      | FRA                | Iga Świątek      |
| 2020 | Wimbledon        | GBR                | Naomi Osaka      |
| 2020 | US Open          | USA                | Naomi Osaka      |
| 2021 | Australian Open  | AUS                | Naomi Osaka      |
| 2021 | French Open      | FRA                | Barbora Krejčíková |
| 2021 | Wimbledon        | GBR                | Ashleigh Barty   |
| 2021 | US Open          | USA                | Emma Raducanu    |

**Table 2: Winner Information**

| Winner             | Nationality |
|--------------------|-------------|
| Sofia Kenin        | USA         |
| Iga Świątek        | POL         |
| Naomi Osaka        | JPN         |
| Barbora Krejčíková | CZE         |
| Ashleigh Barty     | AUS         |
| Emma Raducanu      | GBR         |

## Third Normal Form (3NF)

**Table 1: Tournament Winner**

| Year | Tournament       | Winner           |
|------|------------------|------------------|
| 2020 | Australian Open  | Sofia Kenin      |
| 2020 | French Open      | Iga Świątek      |
| 2020 | Wimbledon        | Naomi Osaka      |
| 2020 | US Open          | Naomi Osaka      |
| 2021 | Australian Open  | Naomi Osaka      |
| 2021 | French Open      | Barbora Krejčíková |
| 2021 | Wimbledon        | Ashleigh Barty   |
| 2021 | US Open          | Emma Raducanu    |

**Table 2: Player Information**

| Player             | Nationality |
|--------------------|-------------|
| Sofia Kenin        | USA         |
| Iga Świątek        | POL         |
| Naomi Osaka        | JPN         |
| Barbora Krejčíková | CZE         |
| Ashleigh Barty     | AUS         |
| Emma Raducanu      | GBR         |

**Table 3: Tournament Information**

| Tournament         | Tournament Country |
|--------------------|--------------------|
| Australian Open    | AUS                |
| French Open        | FRA                |
| Wimbledon          | GBR                |
| US Open            | USA                |

## 3NF with ID numbers

**Table 1: Tournament Winner**

| Year | Tournament ID | Player ID |
|------|---------------|-----------|
| 2020 | 1             | 1         |
| 2020 | 2             | 2         |
| 2020 | 3             | 3         |
| 2020 | 4             | 3         |
| 2021 | 1             | 3         |
| 2021 | 2             | 4         |
| 2021 | 3             | 5         |
| 2021 | 4             | 6         |

**Table 2: Player Information**

| Player ID | Player             | Country ID |
|-----------|--------------------|----------------|
| 1         | Sofia Kenin        | 1              |
| 2         | Iga Świątek        | 2              |
| 3         | Naomi Osaka        | 3              |
| 4         | Barbora Krejčíková | 4              |
| 5         | Ashleigh Barty     | 5              |
| 6         | Emma Raducanu      | 6              |

**Table 3: Tournament Information**

| Tournament ID | Tournament        | Country ID |
|---------------|-------------------|-----------------------|
| 1             | Australian Open   | 5                     |
| 2             | French Open       | 2                     |
| 3             | Wimbledon         | 6                     |
| 4             | US Open           | 1                     |

**Table 4: Country Information**

| Country ID | Country Abbreviation | Country Name    |
|------------|----------------------|-----------------|
| 1          | USA                  | United States   |
| 2          | FRA                  | France          |
| 3          | JPN                  | Japan           |
| 4          | CZE                  | Czech Republic  |
| 5          | AUS                  | Australia       |
| 6          | GBR                  | Great Britain   |
| 7          | POL                  | Poland          |

## Summary of Normal Forms

- **1NF**: Data is organized in a table with each cell holding a single value. Repeated data is allowed.
- **2NF**: Eliminates partial dependencies by ensuring each non-key attribute is fully dependent on the primary key (Year + Tournament here). Data is split into separate tables for different entities (Winners and Nationalities).
- **3NF**: Removes transitive dependencies. Attributes depend only on the primary key, and additional tables are created to represent related data independently (Tournament and Tournament Country).

## Practice

| Artist           | Song            | Album                     | Genres                       |
|------------------|-----------------|---------------------------|------------------------------|
| Taylor Swift     | Cardigan        | *Folklore*                 | Indie Folk, Pop               |
| Ariana Grande    | Positions       | *Positions*                | Pop, R&B                      |
| Dua Lipa         | Levitating      | *Future Nostalgia*         | Pop, Disco                    |

## First Normal Form (1NF)

| Artist           | Song            | Album                     | Genre     |
|------------------|-----------------|---------------------------|-----------|
| Taylor Swift     | Cardigan        | *Folklore*                 | Indie Folk|
| Taylor Swift     | Cardigan        | *Folklore*                 | Pop       |
| Ariana Grande    | Positions       | *Positions*                | Pop       |
| Ariana Grande    | Positions       | *Positions*                | R&B       |
| Dua Lipa         | Levitating      | *Future Nostalgia*         | Pop       |
| Dua Lipa         | Levitating      | *Future Nostalgia*         | Disco     |

## Second Normal Form (2NF)

**Table 1: Songs**

| Song ID | Artist           | Song            | Album                     |
|---------|------------------|-----------------|---------------------------|
| 1       | Taylor Swift     | Cardigan        | *Folklore*                 |
| 2       | Ariana Grande    | Positions       | *Positions*                |
| 3       | Dua Lipa         | Levitating      | *Future Nostalgia*         |

**Table 2: Song Genres**

| Song ID | Genre       |
|---------|-------------|
| 1       | Indie Folk  |
| 1       | Pop         |
| 2       | Pop         |
| 2       | R&B         |
| 3       | Pop         |
| 3       | Disco       |

## Third Normal Form (3NF)

**Table 1: Songs**

| Song ID | Song            | Artist ID | Album ID |
|---------|-----------------|-----------|----------|
| 1       | Cardigan        | 1         | 1        |
| 2       | Positions       | 2         | 2        |
| 3       | Levitating      | 3         | 3        |


**Table 2: Artists**

| Artist ID | Artist           |
|-----------|------------------|
| 1         | Taylor Swift     |
| 2         | Ariana Grande    |
| 3         | Dua Lipa         |


**Table 3: Albums**

| Album ID | Album                     |
|----------|---------------------------|
| 1        | *Folklore*                 |
| 2        | *Positions*                |
| 3        | *Future Nostalgia*         |


**Table 4: Song Genres**

| Song ID | Genre ID  |
|---------|-----------|
| 1       | 1         |
| 1       | 2         |
| 2       | 2         |
| 2       | 3         |
| 3       | 2         |
| 3       | 4         |


**Table 5: Genres**

| Genre ID | Genre       |
|----------|-------------|
| 1        | Indie Folk  |
| 2        | Pop         |
| 3        | R&B         |
| 4        | Disco       |