---
title: "09. Table Joins"
subtitle: "DSST289: Introduction to Data Science"
author: "Erik Fredner"
date: today
highlight-style: arrow
format:
  html:
    anchor-sections: true
    code-tools: true
    df-print: paged
    embed-resources: true
    link-external-icon: true
    link-external-newwindow: true
    number-sections: true
    smooth-scroll: true
    toc: true
editor:
  markdown:
    wrap: 72
---

# Overview

One of the most common tasks in data science is **joining** data from two
(or more) datasets together.

To do this, we need to identify a **key** common between the two datasets.
This key is a feature that uniquely identifies an observation.

If you have never joined data together before, these concepts can be confusing,
so it may be helpful to read this section multiple times.

## Primary Key

A [primary key](https://en.wikipedia.org/wiki/Primary_key) is a feature or group of features (in other words, a column or group of columns) in a dataset that uniquely identifies a row of data.
In our food dataset, the feature `item` is a primary key. In the `us`
dataset, the `city` and `year` columns together form a primary key—we need
both to uniquely identify one observation.

## Foreign Key

Likewise, a [foreign key](https://en.wikipedia.org/wiki/Foreign_key) is the appearance of a primary key within a
different dataset. The `food_group` feature in the `food` dataset, for example,
could be a foreign key if we had another table with one row describing
information about each food group.

## Relation

A primary key and the corresponding foreign key in another table form a
[relation](https://en.wikipedia.org/wiki/Relation_(database)). Typically, a relation maps a single row in one dataset to many rows
in another.

## Table Join

A table join is a way of combining two tables based on relations. The goal is
to match up a foreign key in one table with the primary key in another table,
to add new columns from one dataset into another dataset.

# Examples

## Data

We're going to work with a subset of `food`:

```{r}
#| warning: false

library(tidyverse)

food <- read_csv(file.path("..", "data", "food.csv"))

food |>
  slice_sample(n = 5)
```

## Joining data by key

As an example of performing table joins, we will start with a pared-down
version of our foods data:

```{r}
#| message: false
food_sml <- food |>
  select(item, food_group, calories)

food_sml |>
  slice_sample(n = 5)
```

Within `food_sml`, `item` is a **primary key**.

Now, consider the following new dataset corresponding to dietary restrictions
associated with different food groups:

```{r}
#| message: false
diet <- read_csv(file.path("..", "data", "food_diet_restrictions.csv"))
diet
```

In the `diet` table, `food_group` is a **primary key**.

Within the `food` table, `food_group` is a **foreign key** to `diet`. What we would like to do is to combine these two
datasets by matching up rows that have the same values in the corresponding
columns.

For example, it could be useful to include the columns `vegan`, `vegetarian`,
and `pescatarian` in `food_sml`. To do this, we need to
associate a row in the food dataset with the row in the diet dataset that
contains the same value of the variable `food_group`.

We do this using the
function `left_join`. We pipe in the larger dataset that we want to join columns
to, provide the name of the table to grab columns from (`diet`), and indicate `by` what column the two datasets will be joined.

```{r}
food_sml |>
  left_join(diet, by = "food_group")
```

We now have the dietary restrictions added into the food
dataset. We can now compute models, summaries, and construct visualizations
based on the new metadata associated with each food group.

## Joining a single column

Suppose we *only* wanted to get the `vegan` column from `diet`. We can do this
by selecting the columns we want to keep from the `diet` dataset *before* joining.

There are two different ways of handling this. Both involve what the `tidyverse` style guide refers to as [short pipes](https://style.tidyverse.org/pipes.html#short-pipes).

The first method uses an inline pipe:

```{r}
food_sml |>
  left_join(diet |> select(food_group, vegan), by = "food_group")
```

The second method creates a separate dataset containing only the desired columns, which is then joined:

```{r}
diet_join <- diet |> select(food_group, vegan)

food_sml |>
  left_join(diet_join, by = "food_group")
```

In cases where your pipeline is very short, it is okay to `|>` on a single line
as shown above. However, if you have a longer pipeline, it is better to add
vertical space as we have been doing all semester.

## Using joined data

You can then use the new columns to perform any of the operations we
have already studied. For example, let's figure out which vegan items
have the highest number of calories:

```{r}
food_sml |>
  left_join(diet_join, by = "food_group") |>
  filter(vegan == "yes") |>
  arrange(desc(calories))
```


## Joins with missing matches

In the previous example, all the `food_group` values from `food_sml` were present in `diet`, so all rows were included in the joined table. This is not always the case.

Let's consider a third dataset containing information about food recipes,
one for a Pot Roast and another for Guacamole:

```{r}
#| message: false
recipes <- read_csv(file.path("..", "data", "food_recipes.csv"))
recipes
```

In this dataset, `ingredient` is a foreign key corresponding to the primary key
`item` in the `food_sml` dataset.

For `recipes`, the primary key requires **two**
columns: both `recipe` and `ingredient` are needed to uniquely describe each row.

We can now try adding the `calories` from the `food_sml` dataset into `recipes`. The complication here is that the column we want to join on has
a different name in each dataset ("ingredient" versus "item"). To specify this,
we provide the two different names within the `by` argument of the join function:

```{r}
recipes |>
  left_join(food_sml, by = c("ingredient" = "item"))
```

Now, we have a dataset that has added the food group and calories information
to our recipe dataset.

Notice that the ingredient "Bay Leaf" is not in `food_sml`. Because of this, the food group and calories information for this ingredient are filled in with missing values.

Also, notice that rows about `food` items that are not
in any recipe are *not* included in the output. This is because `left_join` only includes all rows from the left table (`recipes`) and the matching rows from the right table (`food_sml`). Since `food_sml` may contain items not present in recipes, those items are not included in the joined dataset.

# Inner and outer joins

There are four mutating joins available in `dplyr`, all using the same syntax:

- `left_join`: Returns all rows from the first (left) dataset and the matched rows from the second (right) dataset. Unmatched columns from the right dataset are filled with NA.
- `right_join`: Returns all rows from the second (right) dataset and the matched rows from the first (left) dataset. Unmatched columns from the left dataset are filled with NA.
- `full_join`: Returns all rows when there is a match in either left or right dataset. Unmatched columns are filled with NA.
- `inner_join`: Returns only the rows with matches in both datasets.

For example, by replacing `left_join` with the function `inner_join`,
we can modify the join from the last section to drop the missing term
(Bay Leaf) that is not contained in our food dataset:

```{r}
recipes |>
  inner_join(food_sml, by = c("ingredient" = "item"))
```

These four joins are called "mutating" joins because, like `mutate()`,
they *add columns* to a dataset.

There are also two `filter` joins that serve only
to select rows in the first (left) table:

- `semi_join` selects rows in the first table that have a match in the second.
- `anti_join` selects rows in the first table that do not have a match in the second.

These are useful for several tasks and will come up now and then.

# Homework questions

1. Hand-write a dataset that consists of at least two different tables where one table has a primary key that appears as a foreign key in the other table. The tables do not need to be large (10 rows between both tables is more than
enough), and the subject of the tables is up to you. Be creative. Don't
copy the example above.

2. Hand-write a code pipeline that joins the two tables together, then
produces some kind of output (a table, a plot, etc.) using the joined data.

Don't forget to load your handwritten data into R objects. You could do so like this:

```{r}
variable1 <- c("a", "b", "c")
variable2 <- c(1, 2, 3)
data <- tibble(variable1, variable2)
data
```

Note that `variable1` and `variable2` become the column names.