---
title: "Exam 3 Closed Notes"
subtitle: "DSST 289: Introduction to Data Science"
author: "Erik Fredner"
execute:
  warning: false
  message: false
format:
  pdf:
    number-sections: true
    df-print: default
    header-includes: 
      - \usepackage{fontspec}
      - \setmainfont{Atkinson Hyperlegible}
      - \setsansfont{Atkinson Hyperlegible}
      - \setmonofont{IBM Plex Mono}
      - \usepackage{sectsty}
      - \allsectionsfont{\sffamily}
      - \usepackage{setspace}
  html:
    anchor-sections: true
    code-tools: true
    df-print: paged
    embed-resources: true
    link-external-icon: true
    link-external-newwindow: true
    number-sections: true
    smooth-scroll: true
    toc: true
editor:
  markdown:
    wrap: 72
editor_options: 
  chunk_output_type: console
---

```{r}
#| echo: false

library(tidyverse)
library(ggrepel)
library(sf)
theme_set(theme_minimal())
```

# Name

\medskip

Name\hrulefill \medskip

Section start time\hrulefill \medskip

# Exam

- You may only use a pen or pencil and scratch paper on this exam.
- If you cannot solve a problem, write what you do know for partial credit.
- Code will be graded on accuracy and formatting.

# Questions

## Data types

```{r}
#| echo: false
traffic_violations <- tibble(
  violation = c("Speeding", "Illegal parking", "Running red light"),
  occurred = ymd_hms(c(
    "2024-01-01 08:30:00",
    "2024-01-02 14:45:00",
    "2024-01-03 17:20:00"
  )),
  location = st_sfc(
    st_point(c(37.577544, -77.535458)),
    st_point(c(37.578414, -77.534858)),
    st_point(c(37.572642, -77.547447))
  ),
  fine = c(150.0, 75.5, 200.0),
  paid = c(TRUE, FALSE, FALSE)
)
```

```{r}
traffic_violations
```

Identify the data type of each column in `traffic_violations` and describe its purpose. For example: `<int>` represents the integer data type, which stores whole numbers like `1`, as distinct from decimals like `1.5`.

\vspace*{0.2\textheight}

## The novice

Using the table above, a novice programmer is trying to calculate the total amount of unpaid fines. Here is their code so far:

```r
  traffic | > filtar(Paid %is%   FALSE) + mutate[outstanding== sun(fine) ) |>
```

Ultimately, their code should have the following output:

```{r}
#| echo: false
traffic_violations |>
  filter(paid == FALSE) |>
  summarize(outstanding = sum(fine))
```

Identify problems with their code as written. You may either rewrite it correctly *or* list the issues.

\pagebreak

## Average fine

```{r}
#| echo: false
officer_fines <- tibble(
  officer_id = c(101, 102, 101, 102, 101, 102),
  date = ymd(c(
    "2024-01-01", "2024-01-01", "2024-01-02",
    "2024-01-02", "2024-01-01", "2024-01-02"
  )),
  fine = c(100, 150, 200, 100, 300, 400),
  paid = c(TRUE, FALSE, TRUE, TRUE, FALSE, FALSE)
) |> arrange(date)
```

```{r}
officer_fines
```

```{r}
#| eval: false
officer_fines |>
  summarize(average_fine = mean(fine))
```

```{r}
#| eval: false

officer_fines |>
  group_by(officer_id) |>
  summarize(total_fines = sum(fine)) |>
  ungroup() |>
  summarize(average_fine = mean(total_fines))
```

Will `average_fine` differ in the two code blocks above? If so, how?

\vspace*{0.2\textheight}

## Ticket fees

Unpaid tickets automatically incur a late fee. The late fee is one-third of the fine. Using `officer_fines`, write code that will produce the following table:

```{r}
#| echo: false

officer_fines |>
  mutate(
    fee = if_else(paid, 0, fine * (1 / 3)),
    total = fine + fee
  )
```

\vspace*{0.2\textheight}


## Flowers of Virginia

```{r}
#| echo: false

flower <- c(
  "Black-Eyed Susan", "Virginia Bluebell", "Eastern Red Columbine"
)
height_day_1 <- c(10, 8, 6)
height_day_2 <- c(11, 10, 9)
height_day_3 <- c(12, 12, 12)

flowers <- tibble(
  flower,
  height_day_1,
  height_day_2,
  height_day_3
)
```

```{r}
flowers
```

Fill in the blanks to reshape `flowers` such that it can be used to make the plot shown below. *Nota bene*: No custom labels have been applied to the plot.

\pagebreak

```r
_________________ <- flowers |>
  _________________(
    cols = _________________
    names_to = _________________
    names_prefix = "height_day_",
    names_transform = as.integer,
    values_to = _________________
  )
```

```{r}
#| echo: false

flowers |>
  pivot_longer(
    cols = -flower,
    names_to = "day",
    names_prefix = "height_day_",
    names_transform = as.integer,
    values_to = "height"
  ) |>
  ggplot(aes(x = day, y = height, color = flower)) +
  geom_line() +
  geom_point(size = 3) +
  scale_color_viridis_d() +
  scale_x_continuous(breaks = 1:3)
```

## Flower plot

Beginning from your reshaped `flowers`, write code to reproduce the plot above.

\vspace*{0.2\textheight}

## Fast cars

These tables show when and where two artists performed the song "Fast Car:"

```{r}
#| echo: false
setlists <- tibble(
  artist = c(
    "Tracy Chapman", "Tracy Chapman", "Tracy Chapman",
    "Luke Combs", "Luke Combs", "Luke Combs"
  ),
  date = as_date(c(
    "1988-08-01", "1990-07-05", "1995-09-10",
    "2023-02-15", "2023-06-10", "2023-11-20"
  )),
  song = "Fast Car",
  set_order = c(1, 18, 18, 6, 1, 10)
) |>
  arrange(date)
```

```{r}
setlists
```

```{r}
#| echo: false
concert_venues <- tibble(
  artist = c(
    "Tracy Chapman", "Tracy Chapman", "Tracy Chapman",
    "Luke Combs", "Luke Combs", "Luke Combs"
  ),
  date = as_date(c(
    "1988-08-01", "1990-07-05", "1995-09-10",
    "2023-02-15", "2023-06-10", "2024-07-20"
  )),
  venue = c(
    "The Greek", "SummerStage",
    "Red Rocks", "The Greek",
    "Bridgestone Arena", "Ford Field"
  ),
  audience_size = c(6000, 15000, 9000, 6000, 19000, 45000)
)
```

```{r}
concert_venues
```

Write code that will output the following table. After piping from the table you begin with, you may only use *one* function.

```{r}
#| echo: false

concert_venues |>
  inner_join(setlists, by = c("artist", "date"))
```

\pagebreak

## Coordinating changes

```{r}
#| echo: false
#| fig-width: 5
#| fig-cap: "First Virginia map"

states <- read_sf("../data/state.geojson")

states |>
  filter(name == "Virginia") |>
  st_transform(crs = 4326) |>
  ggplot() +
  geom_sf()
```

```{r}
#| echo: false
#| fig-width: 5
#| fig-cap: "Second Virginia map"

states |>
  filter(name == "Virginia") |>
  st_transform(crs = 5069) |>
  ggplot() +
  geom_sf()
```

These plots use the same geospatial data. Only one line of the code has been changed between them. What was it? Why does that change matter for mapmaking? You do not need to write the *exact* line of code, but you do need to explain its function.

\vspace*{0.1\textheight}

## Interpret residuals

```{r}
#| echo: false

library(broom)

hans <- read_csv("../data/hans_roslin.csv")

hans_subset <- hans |>
  filter(year == 2007, continent == "Asia") |>
  select(country, year, gdp, life_exp)

model <- lm(life_exp ~ gdp, data = hans_subset)

hans_augmented <- model |>
  augment(newdata = hans_subset) |>
  arrange(.resid)

countries <- c(
  "Mongolia", "Iran", "Vietnam", "Afghanistan", "China", "Singapore",
  "Japan", "Korea, Rep."
)

hans_augmented |>
  ggplot(aes(gdp, life_exp)) +
  geom_point(alpha = 0.5) +
  geom_text_repel(
    aes(label = if_else(country %in% countries, country, NA)),
    size = 3
  ) +
  geom_smooth(method = "lm", se = FALSE) +
  scale_x_continuous(labels = scales::dollar_format()) +
  labs(
    x = "GDP per person",
    y = "Life expectancy",
    title = "Life expectancy by GDP in Asian countries (2007)"
  )
```

Among the labeled points, which residual has the largest absolute value? Interpret what the residual means for that point with respect to the linear model.

\vspace*{0.2\textheight}

## Extract values from `<dttm>`

```{r}
#| echo: false

destinys_child <- tibble(
  name = c("Beyoncé Knowles", "Kelly Rowland", "Michelle Williams"),
  birth_year = c(1981, 1981, 1979),
  birth_month = c(9, 2, 7),
  birth_day = c(4, 11, 23),
  birth_hour = c(3, 12, 23),
  birth_minute = c(59, 30, 1),
  birth_second = c(5, 28, 44)
) |>
  mutate(
    birth_datetime = make_datetime(
      birth_year,
      birth_month,
      birth_day,
      birth_hour,
      birth_minute,
      birth_second
    )
  ) |>
  select(name, birth_datetime)
```

```{r}
destinys_child
```

Draw the output of the following code:

```{r}
#| eval: false
destinys_child |>
  mutate(
    birth_year = year(birth_datetime),
    birth_month = month(birth_datetime, label = TRUE, abbr = TRUE),
    birth_hour = hour(birth_datetime)
  ) |>
  select(name, birth_year, birth_month, birth_hour)
```

\vspace*{0.2\textheight}

## Coffee shop

```{r}
#| echo: false

# Define operating hours in New York timezone
start_datetime <- ymd_hms("2024-11-15 06:00:00", tz = "America/New_York")
end_datetime <- ymd_hms("2024-11-15 15:00:00", tz = "America/New_York")

# Create a sequence of times at 15-minute intervals
times <- seq(start_datetime, end_datetime, by = "15 mins")

# Create a dataframe
coffee <- tibble(
  start_time = times
) %>%
  mutate(
    # Extract hour and minute
    hour = hour(start_time),
    minute = minute(start_time),
    
    # Calculate decimal time for easier calculations
    decimal_time = hour + minute / 60
  ) %>%
  # Assign people based on time with gradual transitions
  mutate(
    people = case_when(
      # Early Morning: 6 AM to 8 AM - Gradual Increase from 4 to 30
      decimal_time >= 6 & decimal_time < 8 ~ round(
        4 + ((decimal_time - 6) / 2) * (30 - 4) + rnorm(n(), 0, 2)
      ),
      
      # Morning Peak: 8 AM to 10 AM - High occupancy
      decimal_time >= 8 & decimal_time < 10 ~ round(
        30 + runif(n(), 0, 10)
      ),
      
      # Late Morning: 10 AM to 12 PM - Gradual Decrease from 30 to 20
      decimal_time >= 10 & decimal_time < 12 ~ round(
        30 - ((decimal_time - 10) / 2) * (30 - 20) + rnorm(n(), 0, 2)
      ),
      
      # Noon Peak: 12 PM to 1 PM - Temporary Bump to High Occupancy
      decimal_time >= 12 & decimal_time < 13 ~ round(
        20 + runif(n(), 10, 20)
      ),
      
      # Afternoon Decline: 1 PM to 3 PM - Gradual Decrease from 40 to 10
      decimal_time >= 13 & decimal_time <= 15 ~ round(
        40 - ((decimal_time - 13) / 2) * (40 - 10) + rnorm(n(), 0, 2)
      ),
      
      # Default: Minimum Occupancy
      TRUE ~ 4
    )
  ) %>%
  # Ensure people counts are within the specified range
  mutate(
    people = pmin(pmax(people, 4), 40)
  ) %>%
  # Convert start_time to UTC for consistent storage
  mutate(
    time_utc = with_tz(start_time, "UTC")
  ) %>%
  # Select relevant columns
  select(time_utc, people)
```

This dataset observes the number of people in a Richmond, VA coffee shop in 15 minute intervals over the business day, which runs from 6 AM to 3 PM.

```{r}
coffee |> 
  slice_head(n = 3)
```

Plotting `coffee` gives the following:

```{r}
#| echo: false
coffee |>
  ggplot(aes(time_utc, people)) +
  geom_line()
```

What, if anything, is wrong with this plot? How would you fix it? You do not need to write the *exact* line of code, but you do need to explain its function.

\vspace*{0.2\textheight}

# Honor

> "I pledge that I neither gave nor received unauthorized assistance during the completion of this work."

Signature\hrulefill \medskip
